---
title: "HUSTON Monte Carlo Simulation"
author: "Sara Huston"
output:
  pdf_document:
    number_sections: no
    toc: yes
    toc_depth: 2
header-includes: 
- \usepackage{graphicx}
- \usepackage{float}
---

```{r LoadData, message=FALSE, warning=FALSE, include=FALSE}
# DO NOT SHOW IN FINAL REPORT
options(tinytex.verbose = TRUE)
options(digits = 5) 

#Load Libraries
Libraries <- c("knitr", "readr")
for (p in Libraries) { 
    library(p, character.only = TRUE)
}

opts_chunk$set(fig.align='center',
               external=TRUE,
               echo=TRUE,
               warning=FALSE,
               fig.pos='H', 
               tidy.opts=list(width.cutoff=60),
               tidy=TRUE,
               warning = FALSE
               )
```

# 1: Code that allows one to play the game
## Install and load specific packages required for simulation
```{r, message=FALSE}
#install.packages('gdata')
library(dplyr)
library('gdata')
```

## Function that runs the Monty Hall problem
```{r}
MontyHall <- function(guess = integer, change_guess = Boolean) {
  
  # Randomly sets one door as true and others as false
  doors <- sample(c(rep(TRUE, 1), rep(FALSE, 2)), 3, replace = F)
  indicies <- 1:3

  # Open one of the false doors randomly that is not the given guess or the TRUE winner
  potential_removable_doors <- indicies[-c(which(doors), guess)]
  remove_idx <- resample(potential_removable_doors, 1)
  doors[remove_idx] <- NA
  
  if((change_guess && !doors[guess]) || (!change_guess && doors[guess])){
    # Winner
    return (1)
  }
  else{
    # Loser
    return (0)}
}
```

# 2) Simulation in which the player chooses to always keep the original door
choice
```{r}
num_iterations <- 10000
i <- 0
score_nochange <- 0

while (i < num_iterations) {
  guess <- sample(1:3, 1)
  score_nochange = score_nochange + MontyHall(guess, FALSE)
  i = i + 1
}
```

# 3) Simulation in which the player chooses to always switch the door-choice
```{r}
i <- 0
score_change <- 0

while (i < num_iterations) {
  guess <- sample(1:3, 1)
  score_change = score_change + MontyHall(guess, TRUE)
  i = i + 1
}
```

# 4 Compare the effectiveness of the two strategies by observing the outcome of the
simulations
## 4a) From the 10,000 iterations, how many times the player wins the car when
s/he keeps the original door-choice?
```{r}
score_nochange
```

## 4b) From the 10,000 iterations, how many times the player wins the car when
s/he switches the door-choice?
```{r}
score_change
```

## 4c) Based on 4a and 4b, compare the probability of winning the car using each
strategy.
```{r}
print("The percent chance of winning by keeping your original guess:")
score_nochange / num_iterations * 100
print("The percent chance of winning by changing your original guess:")
score_change / num_iterations * 100
```
## The best strategy is changing your guess. This statement is based on the clear difference in results from the simluation data. 



