---
title: "Plots"
author: "Sara Huston"
output:
  pdf_document:
    number_sections: no
    toc: yes
    toc_depth: 2
header-includes: 
- \usepackage{graphicx}
- \usepackage{float}
---

# Assignment 1

## Data Manipulation

```{r LoadData, message=FALSE, warning=FALSE, include=FALSE}
# DO NOT SHOW IN FINAL REPORT
options(tinytex.verbose = TRUE)
options(digits = 5) 

#Load Libraries
Libraries <- c("knitr", "readr")
for (p in Libraries) { 
    library(p, character.only = TRUE)
}

opts_chunk$set(fig.align='center',
               external=TRUE,
               echo=TRUE,
               warning=FALSE,
               fig.pos='H', 
               tidy.opts=list(width.cutoff=60),
               tidy=TRUE,
               warning = FALSE
               )
```

### Install and load all required packages (Don't install in rmd)

```{r, message=FALSE}
# install.packages("dplyr")
# install.packages("tidyverse")
# install.packages("gridExtra")
# install.packages("ggcorrplot")
# install.packages("formatR")

library("ggcorrplot")
library("gridExtra")
library("tidyverse")
library("dplyr")
library("formatR")
```

### Read in files based on type, headers, and with specific NA values accounted for

```{r}
setwd("./Data")
w1_child <- read.csv("w1_child.csv",
                     header = TRUE, na.strings = c("9", "8", "98", "99"))
w2_child <- read.table("w2_child.dat",
                       header = TRUE, , na.strings = c("-999"))
w3_child <- read.csv("w3_child.csv",
                     header = TRUE, na.strings = c("9", "8", "98", "99"))
w4_child <- read.csv("w4_child.csv",
                     header = FALSE, na.strings = c("9", "8", "98", "99"))
names_w4_child <- read.table("names_w4_child.txt")
educinc <- read.csv("educinc.csv",
                    header = TRUE, na.strings = c("."))
```

### Add transposed variable names to w4_child based off of the provided text file

```{r}
names(w4_child) <- t(names_w4_child)
```

### Make all headers into lower case for easier merging

```{r}
names(w1_child) <- tolower(names(w1_child))
names(w2_child) <- tolower(names(w2_child))
names(w3_child) <- tolower(names(w3_child))
names(w4_child) <- tolower(names(w4_child))
names(educinc) <- tolower(names(educinc))
```

### Merge files by famid and select specified variables

```{r, warning=FALSE}
w1234 <- (list(w1_child, w2_child, w3_child, w4_child, educinc)
          %>% reduce(full_join, by = "famid")) %>% dplyr::select(famid, c01cohort, 
          c01gender, c01school, c01sibli, contains("atts"), contains("pcmp"),
          contains("attt"), contains("dscr"), contains("atod"), fameduc, income,
          c01sibli, contains("edex"))
```

### Reverse code for pcmp 1 and 2

```{r}
pcmp_01_02_cols <- c(grep("pcmp01", names(w1234)), grep("pcmp02", names(w1234)))
w1234[ ,pcmp_01_02_cols] <- 5 - w1234[ ,pcmp_01_02_cols]
```

### Compute averages based on sets of columns for variable sets and place average into a new variable

```{r}
w1234$c01attt <- rowMeans(w1234[c(grep("c01attt", names(w1234)))], na.rm = TRUE)
w1234$c04attt <- rowMeans(w1234[c(grep("c04attt", names(w1234)))], na.rm = TRUE)

w1234$c01pcmp <- rowMeans(w1234[c(grep("c01pcmp", names(w1234)))], na.rm = TRUE)
w1234$c04pcmp <- rowMeans(w1234[c(grep("c04pcmp", names(w1234)))], na.rm = TRUE)

w1234$c01dscr <- rowMeans(w1234[c("c01dscr07", "c01dscr08", "c01dscr09", "c01dscr10")], na.rm = TRUE)
w1234$c04dscr <- rowMeans(w1234[c("c04dscr07", "c04dscr08", "c04dscr09", "c04dscr10")], na.rm = TRUE)

w1234$c01atts <- rowMeans(w1234[c("c01atts03", "c01atts07", "c01atts08", "c01atts10")], na.rm = TRUE)
w1234$c02atts <- rowMeans(w1234[c("c02atts03", "c02atts07", "c02atts08", "c02atts10")], na.rm = TRUE)
w1234$c03atts <- rowMeans(w1234[c("c03atts03", "c03atts07", "c03atts08", "c03atts10")], na.rm = TRUE)
w1234$c04atts <- rowMeans(w1234[c("c01atts03", "c04atts07", "c04atts08", "c04atts10")], na.rm = TRUE)

w1234$c01atod <- rowSums(w1234[c("c01atod01", "c01atod02", "c01atod03", "c01atod04", "c01atod05", "c01atod06", "c01atod07", "c01atod08", "c01atod09")], na.rm = TRUE)
w1234$c04atod <- rowSums(w1234[c("c04atod01", "c04atod02", "c04atod03", "c04atod04", "c04atod05", "c04atod06", "c04atod07", "c04atod08", "c04atod09")], na.rm = TRUE)
```

## Quiz Questions

### 1) Dimension of your final data frame (row x column)

```{r}
dim(w1234)
```

### 2) What is the max average education level obtained by parents (“fameduc”)?

```{r}
max_educ <- max(w1234$fameduc, na.rm = TRUE)
max_educ
```

### 3) How many children have parents with this level of education?

```{r}
sum(w1234$fameduc >= max_educ, na.rm = TRUE)
```

### 4) What is the mean for variable “c01attt” for those who have 4 siblings?

```{r}
four_sibs <- w1234[w1234$c01sibli == 4 & !is.na(w1234$c01sibli), ]
mean(four_sibs$c01attt, na.rm = T)
```

### 5) Create difference scores between Waves 1 and 4 for all scales. Which scale has the greatest average absolute change (i.e., difference)?

```{r}
score_variables <- c("atts", "pcmp", "attt", "dscr", "atod")
w1234[paste("difference_", score_variables, sep = "")] <- w1234[paste("c01", score_variables, sep = "")] - w1234[paste("c04", score_variables, sep = "")]
ave_abs_change <- abs(colMeans(w1234[ , c("difference_atts", "difference_pcmp", "difference_attt", "difference_dscr", "difference_atod")], na.rm = TRUE))
ave_abs_change
ave_abs_change[which.max(ave_abs_change)]

abs(mean(w1234$c01atts - w1234$c04atts, na.rm = TRUE))
abs(mean(w1234$c01pcmp - w1234$c04pcmp, na.rm = TRUE))
abs(mean(w1234$c01attt - w1234$c04attt, na.rm = TRUE))
abs(mean(w1234$c01dscr - w1234$c04dscr, na.rm = TRUE))
abs(mean(w1234$c01atod - w1234$c04atod, na.rm = TRUE))
```

# Assignment 2

## Plots

### Plot 1

#### Recode gender variables

```{r}
w1234$gender_r[w1234$c01gender == 1] <- "male"
w1234$gender_r[w1234$c01gender == 2] <- "female"
```

#### Create a new variable called "newedu" w/ various levels of education

```{r}
w1234$new_edu[w1234$fameduc < 7 & !is.na(w1234$fameduc)] <- "Elementary"
w1234$new_edu[w1234$fameduc >= 7 & w1234$fameduc < 13 & !is.na(w1234$fameduc)] <- "High School"
w1234$new_edu[w1234$fameduc >= 13 & w1234$fameduc < 17 & !is.na(w1234$fameduc)] <- "College"
w1234$new_edu[w1234$fameduc >= 17 & !is.na(w1234$fameduc)] <- "More than College"
```

#### Order the education levels and factor

```{r}
w1234$new_edu <- factor(c(w1234$new_edu), levels = c("Elementary", "High School", "College", "More than College"), exclude = NA)
```

#### Remove na from each plot

```{r}
w1234_clean <- subset(w1234, !is.na(new_edu) & !is.na(c04edex01) & !is.na(fameduc))
```

#### Create a box plot with education levels and educational experiences at wave 4

```{r}
g <- ggplot(data = w1234_clean, mapping = aes(new_edu, c04edex01))
g + geom_boxplot(data = w1234_clean, mapping = aes(new_edu, c04edex01), varwidth=T, 
  fill = 'lightblue') + labs(title = "Education Expectations Against Parental Education", 
  x = "Education", y ="Educational Expectations at Wave 4")

```

### Plot 2

#### Create a scatter plot with educational expectations at wave 4 (c04edex01) on the Y-axis with “fameduc” on the X axis

```{r}
ggplot(data = w1234_clean, mapping = aes(x = fameduc, y = c04edex01, color = gender_r, shape = gender_r)) + geom_point() + 
  labs(title = "Children’s Educational Expectations by Family Education with gender.",
       x = "Family Education", y = "Educational Expectations at Wave 4") + 
  scale_shape_discrete(name ="Gender", breaks = c("female", "male"), labels = c("Woman", "Man")) + scale_colour_discrete(name ="Gender", breaks = c("female", "male"), labels = c("Woman", "Man"))

```

### Plot 3

#### Plot histograms of perceived discrimination and use of alcohol, tobacco, or other drugs at waves 1 and 4

```{r}
t1 <- ggplot(data = subset(w1234, !is.na(c01atod)), mapping = aes(c01atod))
t2 <- ggplot(data = subset(w1234, !is.na(c04atod)), mapping = aes(c04atod))
t3 <- ggplot(data = subset(w1234, !is.na(c01dscr)), mapping = aes(c01dscr))
t4 <- ggplot(data = subset(w1234, !is.na(c04dscr)), mapping = aes(c04dscr))

g1 <- t1 + geom_histogram(fill = "lightblue", binwidth = 0.8)+ labs(x = "Drug usage at Wave 1")
g2 <- t2 + geom_histogram(fill = "skyblue", binwidth = 0.7)+ labs(x = "Drug usage at Wave 4")
g3 <- t3 + geom_histogram(fill = "mediumblue", binwidth = 0.2)+ labs(x = "Perceived discrimination at Wave 1")
g4 <- t4 + geom_histogram(fill = "navyblue", binwidth = 0.3)+ labs(x = "Perceived discrimination at Wave 4")

grid.arrange(g1, g2, g3, g4, ncol = 2, nrow = 2)
```

### Describe the distributions in two sentences:

-   Perceived discrimination decreases from wave 1 to wave 4, with an increase of \~100 in 1 scores.
-   Alcohol, tobacco, or other drug use increased heavily from wave 1 to wave 4.

### Plot 4

#### Create a correlogram for all 5 scales at Wave 4

```{r}
cor_scales <- cor(w1234[, c("c04attt", "c04atod", "c04dscr", "c04atts", "c04pcmp")], use = "complete.obs")
cor_scales <- round(cor_scales, 2)
ggcorrplot(cor_scales) + labs(title = "Correlogram for all scales at wave 4") +
  scale_fill_gradient2(name="Pearson\nCorrelation")
```

#### Report in words which variables seem to be correlated.

-   Child’s peer competence (pcmp) seems to be correlated with child’s attachment to teachers (attt) and child’s attachment to school (atts). Child's attachment to teachers (attt) and child's attachement to school (atts) appear also correlated.

## Analysis Questions

### 1) Conduct 2 tests of mean differences

```{r}
hs_group <- subset(w1234, new_edu == "High School" & !is.na(c04edex01))
college_group <- subset(w1234, new_edu == "College" & !is.na(c04edex01))
t.test(hs_group$c04edex01, college_group$c04edex01, parid = TRUE)
t.test(w1234$c01atod, w1234$c04atod, paird = TRUE)
```

#### There was a significant difference in educational expectations at wave 4 between those whose parents have a high school education and those whos parents have a college education. There was a significant difference on alcohol, tobacco, and other drug use between wave 1 and wave 4.

### 2) Two tests for significant correlations

```{r}
cor.test(w1234$c04attt, w1234$c04pcmp, na.action=na.omit)
cor.test(w1234$c04dscr, w1234$c04atod, na.action=na.omit)
```

#### There was a significant difference at wave 4 between child’s peer competence and child’s attachment to teachers. There was not a significant difference at wave 4 between child's alcohol, tobacco, and other drug use and child’s perceived discrimination.
